A Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-02-08 03-semantics.tex
Provas informais no fim da Secao 3.

The $\OUT$ rules \textbf{push} and \textbf{pop} are non-recursive definitions
that only apply to the program as a whole and manipulate the stack level as
follows:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    { \DS e \neq \epsilon }
%   -----------------------------------------------------------
    { \DS \LL p,n,e \RR \OUT \LL bcast(p),n+1,\epsilon \RR }
    & \textbf{(push)}   \\
%%%
& \frac
    { \DS n>0, \2 ((p=@nop) \wedge isblocked(n,p)) }
%   -----------------------------------------------------------
    { \DS \LL p,n,\epsilon \RR \OUT \LL p,n-1,\epsilon \RR }
    & \textbf{(pop)}    \\
%%%
%& \LL p,0,\epsilon \RR \1\xrightarrow\1 \bot
    %& \textbf{(end)}    %\\
\end{eqnarray*}
}
%
Rule \textbf{push} matches whenever there is an emitted event in the triple,
and instantly broadcasts the event to the program, which means
    (a) awaking active \code{await} expressions altogether (see $bcast(p)$ in
        Figure~\ref{fig.bcast}),
    (b) creating a nested reaction by increasing the stack level,
    (c) and, at the same time, consuming the event ($e$ becomes $\epsilon$).
%
Note that rule \textbf{push} is the only one in the semantics that matches an
emitted event and also immediately consumes it.
%
The rule \textbf{pop} only decreases the stack level, not affecting the
program, and does not apply if the program is not blocked (see $isblocked(n,p)$
in Figure~\ref{fig.isblocked}).
This condition ensures that an \code{emit} only resumes after its internal
reaction completes, as discussed in Section~\ref{sec.ceu.evts}.

@@

>>>>

The $\OUT$ rules \textbf{push} and \textbf{pop} are non-recursive definitions
that only apply to the program as a whole and manipulate the stack level as
follows:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    { \DS e \neq \epsilon }
%   -----------------------------------------------------------
    { \DS \LL p,n,e \RR \OUT \LL bcast(p),n+1,\epsilon \RR }
    & \textbf{(push)}   \\
%%%
& \frac
    { \DS n>0, \2 ((p=@nop) \vee isblocked(n,p)) }
%   -----------------------------------------------------------
    { \DS \LL p,n,\epsilon \RR \OUT \LL p,n-1,\epsilon \RR }
    & \textbf{(pop)}    \\
%%%
%& \LL p,0,\epsilon \RR \1\xrightarrow\1 \bot
    %& \textbf{(end)}    %\\
\end{eqnarray*}
}
%
Rule \textbf{push} matches whenever there is an emitted event in the triple,
and instantly broadcasts the event to the program, which means
    (a) awaking active \code{await} expressions altogether (see $bcast(p)$ in
        Figure~\ref{fig.bcast}),
    (b) creating a nested reaction by increasing the stack level,
    (c) and, at the same time, consuming the event ($e$ becomes $\epsilon$).
%
Note that rule \textbf{push} is the only one in the semantics that matches an
emitted event and also immediately consumes it.
%
The rule \textbf{pop} only decreases the stack level, not affecting the
program, and does not apply if the program is not blocked (see $isblocked(n,p)$
in Figure~\ref{fig.isblocked}).
This condition ensures that an \code{emit} only resumes after its internal
reaction completes, as discussed in Section~\ref{sec.ceu.evts}.

@@

\subsection{Determinism, Termination, and Memory Bounds}

- informal discussion

\subsubsection*{Determinism}

The proof for determinism relies on the fact all semantic rules are mutually
exclusive, i.e., their preconditions are unique in the set of rules.
This can be verified by simple rule inspections.

Rule \textbf{push} is the only one with $e \neq \epsilon$ as a precondition,
and is trivially mutually exclusive with all other rules.

Rule \textbf{pop} either has $p=@nop$ or $isblocked(n,p)$ as preconditions.
%
Note that rule \textbf{pop} only syntactically applies to top-level
transitions.
For instance, it can never match $\NST$ rules for subprograms as in rule
\textbf{seq-adv}.
%
Hence, for the first case, rule \textbf{pop} only applies, and is the only one
to apply, to $nop$ as the whole program (i.e., a $nop$ not surrounded by other
expressions, such as in rule \textbf{seq-nop}).
%
For the second case, we need to show that given $\LL p,n,\epsilon \RR$, no
$\NST$ transitions apply with $isblocked(n,p)$ and vice versa.
Except for $@canrun$, there are no $\NST$ transitions for the other blocking
expressions ($awaitExt$, $awaitInt$, $every$, and $fin$).
However, considering the precondition $\LL p,n,\epsilon \RR$,
$isblocked(n,@canrun(n))$ is false.
Hence, given the preconditions for rule \textbf{pop}, no $\NST$ transitions can
occur.
Conversely, if a $\NST$ transition is possible, then $isblocked(n,p)$ must be
false.
Again, except for $@canrun$, all other transitions do not involve blocking
expressions, hence, for these transitions, $isblocked(n,p)$ must be false.
For rule \textbf{canrun}, a transition can only occur if the current stack
level matches $@canrun(n)$.
In this case, $isblocked(n,@canrun(n))$ is false.

Finally, we need to show that $\NST$ transitions are mutually exclusive among
themselves.
%
Note that most rules have unique syntactic prefixes, e.g., $(@nop~@and~q)$
(rule \textbf{and-nop1}) is trivially mutually exclusive with $(@nop~@loop~p)$
(rule \textbf{or-nop1}).
%
The only exceptions are rules \textbf{and-adv1} vs. \textbf{and-adv2}, and
\textbf{or-adv1} vs. \textbf{or-adv2}.
In both cases, we need to show that if the left branch can advance, then it
cannot be blocked and vice-versa, i.e., that 
$\LL p,n,\epsilon \RR \NST \LL p',n,e \RR$ and $isblocked(n,p)$ are mutually
exclusive.
This is the same reasoning for rule \textbf{pop} above.

\subsubsection*{Termination}

- there is always a possible transition until n=0

every cannot restart itself
    - break disallowed
    - emit ignored

\subsubsection*{Memory Bounds}

- program is finite
- lexical scope
    - no heap allocation
- no code reentrancy
    - reexecution only due to loops
    - loop reuse nested vars
-
